 

### 2021-08-18 vertical-thermal-review.Rmd
Lawren's edits

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) the size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves can be cooler than the air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$ (poikilothermy), it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorological conditions (Fig. `r fig_leaf_T`).**
Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling. Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH conditions in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig. `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already be occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as because small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) the size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves can be cooler than the air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ varies with micrometeorological conditions (Fig. `r fig_leaf_T`).**
Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling. Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig. `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. A study of tropical oaks found decreasing stomatal conductive capacity (based on morphology) with increasing site mean $T_{air}$ (Lin et al., 2021), a response likely linked to water conservation under increasing VPD. The extent to which maximum conductance (and its leaf cooling potential) increases with light may therefore depend on the specific thermal microenvironments arrayed across the profile
